---
title: "Investigating Alternate Models"
author: "Ian Dinwoodie"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(car)
knitr::opts_chunk$set(echo=TRUE)
```

# Workspace Setup

## Loading the Data

Load the raw data and verify its dimensions and structure.

```{r}
df <- readRDS('../data/tidy.Rds')
dim(df)
summary(df)
```

## Preparing Common Variables and Functions

```{r}
outcomes <- c(
  'aggression',
  'fear_anxiety',
  'jumping',
  'barking',
  'coprophagia',
  'compulsion',
  'house_soiling',
  'rep_materials',
  'hyperactive',
  'destructive',
  'escape',
  'mounting'
)
outcomes <- sort(outcomes)
```

```{r}
apply_min_xtab <- function(df, outcome, cutoff=10)
{
  drops <- NULL
  for (col in names(df)) {
    if (col == outcome) next
    if (is.integer(df[,col])) next
    
    xtab <- table(df[,col], df[,outcome])
    if (min(xtab) < cutoff) {
      drops <- c(drops, col)
      break
    }
  }
  
  if (length(drops) > 0) {
    cat('\nDropped from model due to insufficient responses:\n')
    cat(drops)
    cat('\n')
  }
  
  return(df[, !(names(df) %in% drops)])
}
```

```{r}
df_exp <- df %>%
  filter(train_6mo_or_less == TRUE)
summary(df_exp)
```

# Alternate Models

## Exploring Training Methods and Equipment

With regard to training methods, the following question was presented to
participants.

> At puppy training classes, what training techniques were used?
> - Rewarding techniques (e.g., treats, praise, pets)
> - Tough love techniques (e.g., yelling, bopping on the nose, swatting on the rump, alpha rolls (pinning on back until dog submits), use of aversive collars (e.g., shock, prong, choke), jerking on the leash, water spraying, scruffing)
> - A combination of rewarding and tough love techniques

However, after discussing with clinicians, the unanimous decision was that all
training methods that involved some form of punishment could be considered
punishment-based. This modification can be seen in the first notebook for this
study (`0_tidy.Rmd`) and the result can be seen by looking at the
`train_technique` column.

```{r}
summary(df_exp$train_technique)
```

In addition, we also presented the following questions to participants:

> What restraining/training devices were employed?
> - Nylon slip collar
> - Buckle collar
> - Head halter (with nose band)
> - Harness (around chest)
> - Metal "choke" collar
> - Prong collar
> - Martingale collar
> - Electric shock collar
> - No devices were employed

We exclude the "Other" response from the list above for simplicity since the
only actionable submissions were for harnesses. The harness count was updated
accordingly.

These devices generally fall into two caregories: punishing and non-punishing.

> Punishing:
> - Metal "choke" collar
> - Prong collar
> - Martingale collar
> - Electric shock collar
> - Nylon slip collar
> Non-punishing:
> - Buckle collar
> - Head halter (with nose band)
> - Harness (around chest)

We calculated the number of dogs exposed to these punishing devices in our
initial notebook.

```{r}
summary(df_exp$punish_device)
```

### Analysis of Grouped Devices

The `punish_device` variable is not enough for our "grouped" analysis since it
does not account for dogs being exposed to no devices. Therefore, we create a
device group (`device_group`) column with three values: punish, non-punish,
none.

```{r}
df_exp <- df_exp %>%
  mutate(device_group <- as.factor(ifelse(
    device_used == FALSE, 'none', ifelse(
      punish_device == TRUE, 'punish', 'non_punish'))))

summary(df_exp$device_group)
```