---
title: "Investigating Alternate Models"
author: "Ian Dinwoodie"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(car)
knitr::opts_chunk$set(echo=TRUE)
```

# Workspace Setup

## Loading the Data

Load the raw data and verify its dimensions and structure.

```{r}
df <- readRDS('../data/tidy.Rds')
dim(df)
summary(df)
```

## Preparing Common Variables and Functions

```{r}
outcomes <- c(
  'aggression',
  'fear_anxiety',
  'jumping',
  'barking',
  'coprophagia',
  'compulsion',
  'house_soiling',
  'rep_materials',
  'hyperactive',
  'destructive',
  'escape',
  'mounting'
)
outcomes <- sort(outcomes)
```

```{r}
apply_min_xtab <- function(df, outcome, cutoff=10)
{
  drops <- NULL
  for (col in names(df)) {
    if (col == outcome) next
    if (is.integer(df[,col])) next
    
    xtab <- table(df[,col], df[,outcome])
    if (min(xtab) < cutoff) {
      drops <- c(drops, col)
      break
    }
  }
  
  if (length(drops) > 0) {
    cat('\nDropped from model due to insufficient responses:\n')
    cat(drops)
    cat('\n')
  }
  
  return(df[, !(names(df) %in% drops)])
}
```

```{r}
df_exp <- df %>%
  filter(train_6mo_or_less == TRUE)
summary(df_exp)
```

# Alternate Models

## Exploring Training Methods and Equipment

With regard to training methods, the following question was presented to
participants.

> At puppy training classes, what training techniques were used?
> - Rewarding techniques (e.g., treats, praise, pets)
> - Tough love techniques (e.g., yelling, bopping on the nose, swatting on the rump, alpha rolls (pinning on back until dog submits), use of aversive collars (e.g., shock, prong, choke), jerking on the leash, water spraying, scruffing)
> - A combination of rewarding and tough love techniques

However, after discussing with clinicians, the unanimous decision was that all
training methods that involved some form of punishment could be considered
punishment-based. This modification can be seen in the first notebook for this
study (`0_tidy.Rmd`) and the result can be seen by looking at the
`train_technique` column.

```{r}
summary(df_exp$train_technique)
```


